---
title: "History of federal elections in Toronto"
author: "PsephoAnalytics"
date: '`r Sys.Date()`'
output:
  html_document:
    css: psepho_styles.css
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(toVotes)
library(toCensus)
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(nlme, quietly = TRUE)
library(ape, quietly = TRUE)
```
```{r data, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
fed_votes_2006_geo@data$year <- 2006
fed_votes_2008_geo@data$year <- 2008
fed_votes_2011_geo@data$year <- 2011
fed_votes <- toFederalVotes
# Summarize just the three major parties, the rest are "Other"
major_parties <- c("Conservative", "Liberal", "NDP")
# Standardize the names
levels(fed_votes$party)[c(16, 17, 19)] <- c("NDP", "NDP", "Conservative")
levels(fed_votes$party)[!(levels(fed_votes$party) %in% major_parties)] <- "Other"
droplevels(fed_votes)
# Summarize votes by year, district, and poll
votes <- fed_votes %>%
  mutate(# Clean up polling labels
    # -[letter]|[number] indicate sub-polls and should be merged
    poll = stringr::str_replace(poll, "(\\d{1,3})(-\\d+\\w?|\\D$)", "\\1")
  ) %>%
  group_by(year, district, poll, party) %>%
  summarize(votes = sum(votes)) %>%
  mutate(prop_votes = votes/sum(votes))
```

# Historical patterns

We start by looking at how votes per party have changed over time. Since the polling boundaries and locations change each year, we've standardized the geography of voting by aggregating poll boundaries into census tracts.

```{r votes_by_year, echo=FALSE, message=FALSE, cache=TRUE}
ct_vote_plot <- function(geo_data, year) {
  library(dplyr)
  data <- geo_data@data %>%
    tidyr::gather(party, votes, -id, -year) %>%
    dplyr::group_by(id, party, add = FALSE) %>%
    dplyr::summarize(votes = sum(votes, na.rm = TRUE)) %>%
    dplyr::mutate(prop_votes = votes/sum(votes)) %>%
    dplyr::select(-votes)
  geo_data@data <- dplyr::select(geo_data@data, id)
  geo <- ggplot2::fortify(geo_data, region="id")
  library(ggplot2)
  ggplot(data, aes(map_id = id)) +
    geom_map(aes(fill = cut_interval(prop_votes, length = 0.15)), map = geo) +
    scale_fill_brewer("Proportion of votes", labels=c("Low", rep("", 4), "High"),
                      type = "seq", palette = "YlOrBr") +
    labs(x="", y="", title=paste0(year, " Federal General Election")) +
    theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), # get rid of x ticks/text
          axis.ticks.x = element_blank(), axis.text.x = element_blank(), # get rid of y ticks/text
          plot.title = element_text(lineheight=.8, face="bold", vjust=1)) + # make title bold and add space +
    expand_limits(x = geo$long, y = geo$lat) +
    facet_wrap(~party, as.table = FALSE)
}
```

```{r historical_plots, echo=FALSE, message=FALSE, cache=TRUE}
ct_vote_plot(fed_votes_2006_geo, 2006)
ct_vote_plot(fed_votes_2008_geo, 2008)
ct_vote_plot(fed_votes_2011_geo, 2011)
```

One pattern that is consistent across all years is the relatively poor showing of any of the "Other" parties, even when considered in the aggregate. Across the years, we can see a remarkable decline in support for the Liberals in the 2011 election, after a relatively strong showing in the 2008 election. NDP support seems to grow from just the downtown in 2006 to a much wider distribution in 2011. Conservative support, conversely, seems to expand in the 2008 election (though not enough to win any seats), to more focused support in specific areas (winning them eight seats in total).

We can take a more direct look at how party support changes over time by plotting the change in the proportion of votes received by each party between 2006 and 2011.

```{r change_in_votes, echo=FALSE, message=FALSE, cache=TRUE}
data <- dplyr::bind_rows(fed_votes_2006_geo@data, 
                         fed_votes_2008_geo@data, 
                         fed_votes_2011_geo@data) %>%
  tidyr::gather(party, votes, Other, Conservative, Liberal, NDP) %>% 
  dplyr::filter(year != 2008) %>% 
  dplyr::group_by(year, id, party) %>%
  dplyr::summarize(votes = sum(votes, na.rm = TRUE)) %>% 
  dplyr::mutate(prop_votes = votes/sum(votes, na.rm = TRUE)) %>% 
  dplyr::select(-votes) %>% 
  dplyr::group_by(id, party) %>% 
  dplyr::summarize(change_votes = diff(prop_votes))
geo <- ggplot2::fortify(fed_votes_2011_geo, region="id")
ggplot(data, aes(map_id = id)) +
  geom_map(aes(fill = cut_interval(change_votes, n = 5)), map = geo) +
  scale_fill_brewer("Change in proportion", labels=c("Decrease", rep("", 3), "Increase"),
                    type = "div", palette = "BrBG") +
  labs(x="", y="", title= "Votes in the 2006 and 2011 Federal General Elections") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), # get rid of x ticks/text
        axis.ticks.x = element_blank(), axis.text.x = element_blank(), # get rid of y ticks/text
        plot.title = element_text(lineheight=.8, face="bold", vjust=1)) + # make title bold and add space
  expand_limits(x = geo$long, y = geo$lat) +
  facet_wrap(~party, as.table = FALSE)
```

Clearly, the story from 2006 to 2011 is a large, widespread drop in Liberal party support (with an average decline of `r round(abs(mean(filter(data, party == "Liberal", !is.nan(change_votes))$change_votes)),2)*100`% and maximum decline of `r round(abs(min(filter(data, party == "Liberal", !is.nan(change_votes))$change_votes)),2)*100`%). The benefits of this decline were roughly equally shared by the Conservative and NDP parties, but in different locations. The NDP seems to have particularly benefited from the Liberal's decline in the north east.

# Role of incumbents and star candidates

Generally, incumbents in elections are expected to do well. Parties also devote considerable resources to promoting "star" candidates. How important are these two effects? The Elections Canada data indicates which candiates were incumbents in each election. For star candiates, we reviewed the list for each of the three elections and determined "star" status based on name recognition. These allow us to investigate how incumbancy and star status affect the proportion of votes received by a candidate. They also allow us to determine how important the effects are relative to each other and whether they can reinforce each other.

We start with a simple plot showing the proportion of votes received for the four types of candidate.

```{r incumbents, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
# Add in star candidate status
star_status <- readr::read_csv("data/star_status.csv") %>%
  # readxl::read_excel("data/2015 Candidates.xlsx", skip = 1)
  # Need to reorganize candidate names to use the Excel file
  mutate(year = as.factor(year),
         candidate = as.factor(candidate))
candidates <- dplyr::left_join(fed_votes, star_status) %>%
  group_by(year, district, poll, party) %>%
  summarize(star_candidate = as.logical(max(star_candidate)),
            incumbent = as.logical(max(incumbent)),
            n = n()) %>%
  select(-n)
data <- dplyr::left_join(candidates, votes)
candidate_model <- lme(prop_votes ~ party + star_candidate*incumbent, data = data, 
                    random = ~1|year/district, na.action = na.exclude)
candidate.effects <- fixed.effects(candidate_model)
ggplot(data, aes(x = factor(incumbent), y = prop_votes, fill = factor(star_candidate))) + 
  geom_boxplot() +
  scale_fill_brewer(palette = 7) +
  scale_x_discrete("Incumbent") +
  scale_y_continuous("Proportion of votes") +
  guides(fill = guide_legend(title = "Star candidate"))
```

A mixed-effects model of these data show that being either a star candidate or an incumbent can boost a candidates share of the vote by `r round(mean(candidate.effects["star_candidateTRUE"], candidate.effects["incumbentTRUE"]), 2)*100`%. But, as is clear in the figure above, being both an incumbent and a star candidate does not give a candidate an incremental increase. In other words, the effects are not additive, as evidenced by the statistically significant interaction between the two effects.

Furthermore, the two effects are equivalent to belonging to a party (`r round(mean(candidate.effects["partyConservative"], candidate.effects["partyLiberal"], candidate.effects["partyNDP"]), 2)*100`%). So, being a member of a major party and either incumbent or a star candidate (i.e., heavily promoted) offers the best chance of winning an election. Perhaps not a surprising finding overall, but the relative equality of the three effects, plus negative interaction of incumbency and star candidate, are useful nuances.

Interpretation needs to be careful though. PCs don't have any incumbents in the data and NDP don't have any non-incumbent stars.

# Party leaders

Another prominant feature of modern elections is the focus on the party leader. We've assigned each party leader a "likeability score" for each election and can include this factor in our analysis of the proportion of votes received.

```{r leaders, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
leaders <- readxl::read_excel("data/Leader likeability.xlsx", sheet = "Historic Likeability", skip = 0) %>% 
  dplyr::select(Year, Party, Leader, Likeability) %>% 
  dplyr::mutate(Party = as.factor(Party),
                Year = as.factor(Year))
names(leaders) <- tolower(names(leaders))
data <- dplyr::left_join(data, leaders) %>% 
  dplyr::filter(party != "Other")
leader_model <- lme(prop_votes ~ party + likeability + star_candidate*incumbent, data = data, 
                    random = ~1|year/district, na.action = na.exclude)
leader_effects <- fixed.effects(leader_model)
# summary(leader_model)
```

A linear, mixed effects model finds that leader likeability is associated with a `r round(leader_effects["likeability"][[1]], 3)*100`% change in the proportion of votes received by a candidate. For example, Stephan Dion caused an average `r (leaders$Likeability[leaders$Leader == "Paul Martin"][[1]] - leaders$Likeability[leaders$Leader == "Stephane Dion"][[1]])*round(leader_effects["likeability"][[1]], 3)*100`% decline for Liberal candidates in 2008, relative to Paul Martin.

# Positions

The final component from a candiates perspective that we consider is the parties' platforms. Here we've analyzed the parties' platforms from previous years and scored them along six topics: economy, environment, foreign policy, government, social, and health and education. 

```{r platforms, echo=FALSE, warning=FALSE, message=FALSE}
platforms <- readxl::read_excel("data/Position Scores.xlsx", sheet = "Summary Postions", skip = 29)[,1:8] %>% 
  dplyr::filter(!is.na(Year))
names(platforms) <-tolower(names(platforms))
names(platforms)[c(2, 5, 8)] <- c("party", "foreign_policy", "health_education")
platforms$year <- as.factor(platforms$year)
platforms$party <- as.factor(platforms$party)
platforms <- tidyr::gather(platforms, topic, score, -year, -party)
platform_data <- dplyr::left_join(votes, platforms) %>% 
  dplyr::filter(party != "Other",
                topic %in% c("environment", "economy"))
platform_model <- lme(prop_votes ~ party + topic, data = platform_data, 
                    random = ~1|year/district, na.action = na.exclude)
platform_spread <- tidyr::spread(platforms, topic, score)
platform_spread <- dplyr::left_join(votes, platform_spread) %>% 
  dplyr::filter(party != "Other")
platform_spread_model <- lme(prop_votes ~ party + economy + 
                               environment + foreign_policy + 
                               government + social, 
                             data = platform_spread, 
                    random = ~1|year/district, na.action = na.exclude)
focused_platform <- update(platform_spread_model, . ~ party * (economy + environment))
platform_effects <- fixed.effects(focused_platform)
```

Considering all `r length(levels(platform_data$topic))` topics, only the economy and the environment have significant effects on the proportion of votes received by the parties. High scores for either the economy or the environment decrease the proportion of votes for the Conservatives. The economic effect is very small (`r round(platform_effects["economy"][[1]],3)`) and the environmental one is as strongly negative (`r round(platform_effects["environment"][[1]],3)`) as the Liberals is positive. Both topics increase votes for the Liberals and the NDP. However, the Liberals benefit `r round(platform_effects["partyLiberal:economy"][[1]]/platform_effects["partyNDP:economy"][[1]], 0)` times as much on economic topics, while the NDP benefit `r round(platform_effects["partyNDP:environment"][[1]]/platform_effects["partyLiberal:environment"][[1]], 1)` times as much on environmental.

# Demographics

Now that we've looked at aspects of candidates, parties, and their leaders, we can move to attributes of the voters themselves. In this case, we use data from the 2011 Census, specifically age, gender, and family income.

```{r demographics, echo=FALSE, message=FALSE, cache=TRUE}
data <- dplyr::bind_rows(fed_votes_2006_geo@data, 
                         fed_votes_2008_geo@data, 
                         fed_votes_2011_geo@data) %>%
  tidyr::gather(party, votes, Other, Conservative, Liberal, NDP) %>% 
  mutate(Geo_Code = as.character(id), 
         year = as.factor(year)) %>% 
  # dplyr::filter(party != "Other") %>% 
  dplyr::group_by(year, Geo_Code, party) %>%
  dplyr::summarize(votes = sum(votes, na.rm = TRUE)) %>% 
  dplyr::mutate(prop_votes = votes/sum(votes, na.rm = TRUE)) # %>% 
demog <- toCensus %>% 
  group_by(Geo_Code) %>% 
  mutate(gender = as.integer(gender)-1) %>% 
  summarize(gender = mean(gender, na.rm = TRUE),
            age = mean(age, na.rm = TRUE),
            family_income = mean(family_income, na.rm = TRUE))
data <- dplyr::left_join(data, demog)
demog_model <- lme(prop_votes ~ (gender + age + family_income) * party, data = data, 
                    random = ~1|year/Geo_Code, na.action = na.exclude)
# summary(demog_model)
demog_effects <- fixed.effects(demog_model)
```

The linear, mixed effects model shows that the relationships between age, gender, family income, and the proportion of votes vary widely across the parties (as expected). In general, the effects are:

- Age tends to increase support for Conservatives and Liberals (though the slope is higher for Conservatives; `r round(demog_effects["age:partyConservative"][[1]], 3)` vs `r round(demog_effects["age:partyLiberal"][[1]], 3)`), while decreasing support for NDP (`r round(demog_effects["age:partyNDP"][[1]], 3)`).
- Family income tends to increase support for Conservatives (`r round(demog_effects["family_income:partyConservative"][[1]]*10000,3)`/$10,000) while decreasing for the other two major parties by roughly the same magnitude.
- Gender is a surprise. We are detecting a strong signal for increased support by women for Conservatives, moderate support for Liberals, and strong negative support for NDP. This is not consistent with the notion that women favour more progressive parties and requires further analysis.

One important point to make is that we are not actually tracking the demographics of voters themselves. Rather, we are using census data for the "neighbourhoods" in which the voters cast their ballots. Consequently, there is a chance that neighbourhoods with a higher proportion of women are voting for Conservative candidates, even if the individual women are more likely to vote for a different party. There is no evidence for this, however, and there is no particular reason to expect that such an effect would only influence Conservative support.

The figure below takes a closer look at the relationship between gender and the proportion of votes received by each party. The year of the election is indicated by the colour.

```{r gender_plot, echo=FALSE, warning=FALSE, cache=TRUE}
ggplot(filter(data, party != "Other"), aes(x = gender, y = prop_votes, colour = year)) + 
  geom_point() +
  scale_colour_brewer(type = "seq", palette = 7) +
  scale_x_continuous("Proportion female") +
  scale_y_continuous("Proportion of votes") +
  guides(colour = guide_legend(title = "Year")) +
  facet_wrap(~party)
```

One evident pattern is that the points are oriented vertically. A strong gender effect would appear as a horizontal, or at least angled, line. For the most part, each party's support is clustered around a relatively narrow band just above 50% female. This may be a limitation of these data. On average, Toronto is about 52% female (according to the census data) and this proportion is roughly the same across census tracts. There may simply not be enough variation in gender at the census tract level to reveal an important relationship with voting.

However, if we look a little closer at the bottom left of the panel for Conservatives, we see a cluster of census tracts with a relatively high proportion of males that has low support for Conservatives (below about 0.2 for proportion of votes). Liberals appear to have support from some of these high-male proportion census tracts between about 0.3 and 0.6 for proportion of votes and the NDP from about 0.1 to 0.5. The gender effect may, in fact, be due to specific, predominantly male census tracts that tend to note favour the Conservatives.

# Putting it all together

We've looked at different aspects of candidates, their leaders, and voters. In the end, all of these things matter when votes are actually cast. So, we now consider all aspects simultaneously.

```{r comprehensive, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
platforms <- readxl::read_excel("data/Position Scores.xlsx", sheet = "Summary Postions", skip = 29)[,1:8] %>% 
  dplyr::filter(!is.na(Year))
names(platforms) <-tolower(names(platforms))
names(platforms)[c(2, 5, 8)] <- c("party", "foreign_policy", "health_education")
platforms$year <- as.factor(platforms$year)
platforms$party <- as.factor(platforms$party)
candidates <- candidate_attr %>% 
  mutate(Geo_Code = id,
         year = as.factor(year)) %>% 
  select(-id)
data <- dplyr::left_join(data, candidates)
data <- dplyr::left_join(data, leaders) %>% 
  dplyr::filter(party != "Other")
data <- dplyr::left_join(data, platforms)
comprehensive_model <- lme(prop_votes ~ party * (gender + age + family_income) +
                             (economy + environment) +
                             likeability + star_candidate*incumbent, 
                           data = data, 
                    random = ~1|year/Geo_Code, na.action = na.exclude)
# summary(comprehensive_model)
comprehensive_effects <- fixed.effects(comprehensive_model)
dplyr::data_frame(coefficient = names(comprehensive_effects), value = comprehensive_effects)
```
Each of these coefficients are statistically signficant components of the model.

# Predictions

We start with a high-level prediction for Toronto, without considering details at the census tract level.

```{r toronto_prediction}
pred_leaders <- readxl::read_excel("data/Leader likeability.xlsx", sheet = "Likeability")[1,c(2,11:13)]
names(pred_leaders) <- c("year", "Conservative", "NDP", "Liberal")
pred_leaders$year <- as.factor(lubridate::year(pred_leaders$year))
pred_leaders <- tidyr::gather(pred_leaders, party, likeability, -year)
pred_platforms <- dplyr::filter(platforms, year == 2015,
                                topic %in% c("environment", "economy")) %>% 
  tidyr::spread(topic, score)
pred_candidates <- readxl::read_excel("data/2015 Candidates.xlsx", skip = 1)[,c(2:5)]
names(pred_candidates) <- c("candidate", "party", "incumbent", "star_candidate")
      %>% 
  # candidates %>% 
pred_candidates <- filter(pred_candidates, party != "Other") %>% 
  group_by(party) %>% 
  summarize(star_candidate = mean(star_candidate),
            incumbent = mean(incumbent))
pred_demog <- demog %>% 
  dplyr::group_by(Geo_Code) %>% 
  dplyr::summarize(year = as.factor(2015),
            gender = mean(gender, na.rm = TRUE),
            age = mean(age, na.rm = TRUE),
            family_income = mean(family_income, na.rm = TRUE))
# pred_geo <- dplyr::data_frame(Geo_Code = unique(data$Geo_Code), year = as.factor(2015))
pred_data <- dplyr::left_join(pred_candidates, dplyr::left_join(pred_platforms, pred_leaders))
pred_data <- dplyr::left_join(pred_data, dplyr::left_join(pred_geo, pred_demog))
pred <- predict(comprehensive_model, pred_data, level = 0, na.action = na.omit)
results <- dplyr::data_frame(prop_votes = pred) %>% 
  dplyr::bind_cols(dplyr::filter(pred_data, !is.na(family_income)))
results %>% 
  dplyr::group_by(party) %>% 
  dplyr::summarise(prop_votes = mean(prop_votes))
```

